---
title: "Machine Learning - Predicting exercise execution type"
author: "Martin Bühler, Feb 15, 2022"
output: html_document
---

# 1 Management summary
In this project we use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants.
The participants were asked to barbell lifts correctly and incorrectly in 5 different ways
and we want to predict in which of the 5 manners the exercise was done.

Because the outcome is a factor variable we look at predicting with trees, random forests and boosting.
The best option turns out to be the random forest algorithm where we can achieve an accuracy of over 0.99
and therefore an expected out of sample error of less than 1%.

We can confirm that with both, the random forest and boost algorithm (with 0.632 Bootstrap for cross validation),
we have the same prediction on the test set.
Which also serves as further cross validation evidence on the accuracy of our model.

# 2 Cleaning of data

## 2.1 Data inspection
First of all, the relevant libraries for this project are loaded,
the training data is stored in the data frame df.training
and the test data is imported in the data frame df.testing.

```{r, results='hide', message=FALSE, warning=FALSE}
# Initialize libraries
library(caret)
library(rattle)
library(randomForest)
library(gbm)
library(ggplot2)

# Extract the data
df.training <- read.csv2("pml-training.csv", sep = ",", header = TRUE)
df.testing <- read.csv2("pml-testing.csv", sep = ",", header = TRUE)
```

## 2.2 Data inspection
When inspecting the data we can observe the following:  
- Some of the columns contain only little data (2% or 406 of 19216 rows are only filled in those columns)  
--> The same columns in the training data set are not filled either and using them for prediction is therefore not useful, thus we can drop them from the training set  
- Also some columns are of type string when containing numbers  
- Further the first seven columns are descriptive and can't be used for prediction  
- The class column isn't a factor column and has to be converted too  

```{r, results='hide'}
# Inspect the data
str(df.training, list.len=ncol(df.training))
colSums(is.na(df.training) | df.training == "")
```

## 2.3 Data cleaning
The following code is implementing the findings as observed in the chapter "2.2 Data inspection".
The resulting data frames keep the same name.

```{r}
# Drop columns with many invalid entries
df.training <- df.training[ ,colSums(is.na(df.training) | df.training == "") == 0]
df.testing <- df.testing[ ,colSums(is.na(df.testing) | df.testing == "") == 0]
# Change data type of string columns to 

# Drop descriptive columns
df.training <- df.training[, 8:ncol(df.training)]
df.testing <- df.testing[, 8:ncol(df.testing)]

# Convert columns to numeric
df.training[, c(1:(ncol(df.training)-1))] <- sapply(df.training[, c(1:(ncol(df.training)-1))], as.numeric)
df.testing[, c(1:(ncol(df.testing)-1))] <- sapply(df.testing[, c(1:(ncol(df.testing)-1))], as.numeric)

# Convert last column to factor column in training set
df.training$classe <- as.factor(df.training$classe)
```

# 3 Machine learning algorithms

## 3.1 Predicting with trees

First we try Predicting with trees, an algorithm which is relatively efficient.

```{r}

ctrl = rpart.control(maxdepth=6, cp=0)
fit <- train(classe ~ . , method="rpart", data=df.training, control=ctrl)
fancyRpartPlot(fit$finalModel)
```
It turns out that in the final model not all factors of the classe variable are present.
The variable D is not showing up in the decision tree
and therefore we have to find another prediction algorithm.  
  
Nevertheless its useful to look at the boxplots to better understand the variables 
that indeed showed up in the decision tree
and we can observe quite a big difference between the classes for some of the variables.  

```{r}
par(mfrow=c(2,2))
plot(df.training[c(ncol(df.training),1)])
plot(df.training[c(ncol(df.training),38)])
plot(df.training[c(ncol(df.training),40)])
plot(df.training[c(ncol(df.training),41)])
```

## 3.2 Random Forests and Boosting 

The following code is training on the Random Forest and Boosting algorithm
and showing their respective output.
```{r}
fitControl <- trainControl(method = "boot632", number = 1, summaryFunction=defaultSummary)
fit.rf <- train(classe ~ ., method="rf", metric='Accuracy', trControl = fitControl, data=df.training)
fit.gbm <- train(classe ~ ., method="gbm", metric='Accuracy', trControl = fitControl, data=df.training, verbose=FALSE)
fit.rf
fit.gbm
```
# 4 Conclusion

From the output of the train function in the caret package we can conclude
that both selected models are very accurate
- Random Forest: > 0.995 accuracy
- Boosting: > 0.966 accuracy

Also when looking at the the prediction of the testing set,
we can confirm that both models agree on the predictions. 

```{r}
identical(predict(fit.rf,df.testing), predict(fit.gbm,df.testing))
```



